Causal inference is not a statistical problem

### 2023-04-05 manuscript.qmd
update to add context for other quartets

expander <- function() {
  expand_plot(
    expand_x = expansion(c(0.01, 0.05)), 
    expand_y = expansion(c(0.2, 0.2))
  )
}

@@

d_mbias %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(4) M-bias") +
  expander() -> p_m_bias

d_coll %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(1) Collider") +
  expander()  -> p_coll

d_med %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(3) Mediator") +
  expander()  -> p_med

d_conf %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(2) Confounder") + 
  expander() -> p_conf

(p_coll + plot_spacer() + p_conf) /  (p_med + plot_spacer() + p_m_bias)  
```

```{r}
#| fig-cap: "100 points generated using the data generating mechanisms specified (1) Collider (2) Confounder (3) Mediator (4) M-Bias. The blue line displays a linear regression fit estimating the relationship between X and Y, in each case the slope is 1. "
#| label: fig-2
library(tidyverse)
# devtools::install_github("LucyMcGowan/quartet")
library(quartet)

ggplot(causal_quartet, aes(x = x, y = y)) +
  geom_point(alpha = 0.25) + 
  geom_smooth(
    method = "lm", 
    formula = "y ~ x", 
    linesize = 1.1, 
    color = "steelblue"
  ) +
  facet_wrap(~dataset)
```

@@

```{r}
#| label: tbl-2
#| tbl-cap: "Coefficients for the exposure under each data generating mechanism depending on the model fit as well as the correlation between $X$ and $Z$."

@@

@@

```{r}
#| eval: false
#| echo: true
#| ref.label: ["fig-2", "tbl-2"]
```

>>>>

@@

d_mbias %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(4) M-bias") -> p_m_bias

d_coll %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(1) Collider") -> p_coll

d_med %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(3) Mediator") -> p_med

d_conf %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(2) Confounder") -> p_conf

(p_coll + p_conf) /  (p_med + p_m_bias)  
```

```{r}
n <- 100
set.seed(1331)

confounding <- tibble(
  z = rnorm(n),
  x = z + rnorm(n),
  y = 0.5 * x + z + rnorm(n)
)

set.seed(229655)

collider <- tibble(
  x = rnorm(n),
  y = x + rnorm(n),
  z = 0.45 * x + 0.77 * y + rnorm(n)
)

set.seed(207052)

mediator <- tibble(
  x = rnorm(n),
  z = x + rnorm(n),
  y = z + rnorm(n)
)

set.seed(839)

m_bias <- tibble(
  u1 = rnorm(n),
  u2 = rnorm(n),
  z = 8 * u1 + u2 + rnorm(n),
  x = u1 + rnorm(n, sd = 1),
  y =  x + u2 + rnorm(n, sd = 1)
)

d <- bind_rows(confounding, collider, mediator, m_bias) %>%
  mutate(type = rep(c("(2) Confounder", "(1) Collider", "(3) Mediator", "(4) M-Bias"), each = 100))
```

```{r}
#| fig-cap: "100 points generated using the data generating mechanisms specified (1) Collider (2) Confounder (3) Mediator (4) M-Bias. The blue line displays a linear regression fit estimating the relationship between X and Y, in each case the slope is 1. "
#| label: fig-2
ggplot(d, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") +
  facet_wrap(~type)
```

@@

```{r}
#| label: tbl-2
#| tbl-cap: "Coefficients for the exposure under each data generating mechanism depending on the model fit as well as the correlation between $X$ and $Z$."
d %>%
  nest_by(type) %>%
  mutate(`Y ~ X` = round(coef(lm(y ~ x, data = data))[2], 2),
         `Y ~ X + Z` = round(coef(lm(y ~ x + z, data = data))[2], 2),
         `Correlation of X and Z` = round(cor(data$x, data$z), 2)) %>%
  select(-data, `Data generating mechanism` = type) %>%
  knitr::kable()

@@

The use of small sets of data to demonstrate a key concept in the spirit of Anscombe's Quartet has been applied to a wide variety of data analytic problems. Recent examples include an extension of the original idea proposed by Anscombe called the "Datasaurus Dozen" [@matejka2017same], an exploration of varying interaction effects [@rohrer2021precise], a quartet of model types fit to the same data that yield the same performance metrics but fit very different underlying mechanisms [@biecek2023performance], and a set of conceptual causal quartets that highlight the impact of treatment heterogeneity on the average treatment effect [@gelman2023causal]. While similar in name, the conceptual causal quartets are different from what we present here as they provide excellent insight into how treatment heterogeneity can impact an average treatment effect (by plotting the latent true causal effect) rather than providing a set of observable data as we have here. We believe both sets provide important and complementary understanding for data analysis practitioners.

@@

```{r}
#| eval: false
#| echo: true
library(tidyverse)
# devtools::install_github("LucyMcGowan/quartet")
library(quartet)

ggplot(causal_quartet, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") +
  facet_wrap(~dataset)
